---
title: "Example Code for Challenges"
author: "Coded by Allison"
date: "October 25, 2019"
output: html_document
---


# Challenge 2 
## Part 3 

What was the heaviest animal measured in each year? Return the columns `year`, 
`genus`, `species_id`, and `weight`.


```{r, answer=TRUE, purl=FALSE}
surveys %>%
  filter(!is.na(weight)) %>%
  ## I needed to remove the NA's 
  group_by(year) %>%
  filter(weight == max(weight)) %>%
  ## This comparison fails, if you have NA's in your variable! 
  select(year, genus, species, weight) %>%
  arrange(year)
```


# Challenge 4  
## Part 1  

Spread the `surveys` data frame with `year` as columns, `plot_id` as rows, and 
the number of genera per plot as the values. You will need to summarize before
reshaping, and use the function `n_distinct()` to get the number of unique
genera within a particular chunk of data. It's a powerful function! See
`?n_distinct` for more.


```{r, answer=TRUE, purl=FALSE}
surveys_spread_genera <- surveys %>%
  group_by(plot_id, year) %>%
  summarize(n_genera = n_distinct(genus)) %>%
  spread(year, n_genera)
 
head(surveys_spread_genera)
```

## Part 2 

Now take that data frame and `gather()` it again, so each row is a unique
`plot_id` by `year` combination.


```{r, answer=TRUE, purl=FALSE}
surveys_spread_genera %>%
  gather("year", "n_genera", -plot_id)
  ## first argment is what the name of the new column (with the gathered columns) will be
  ## second argument is what the name of the value column (with the numbers) will be
  ## third argment is what columns you want to gather (- means everything but this column)
```


## Part 3 

The `surveys` data set has two measurement columns: `hindfoot_length` and 
`weight`.  This makes it difficult to do things like look at the relationship 
between mean values of each measurement per year in different plot types. Let's
walk through a common solution for this type of problem. First, use `gather()` 
to create a dataset where we have a key column called `measurement` and a
`value` column that takes on the value of either `hindfoot_length` or
`weight`. *Hint*: You'll need to specify which columns are being gathered.


```{r, answer=TRUE, purl=FALSE}
surveys_long <- surveys %>%
   gather(key = "measurement", value = "value", hindfoot_length, weight)
   ## gathers the hindfoot_length and weight columns into one new column called
   ## measurement and places the values in these columns in to a new column
   ## called value 

## Remember: when you gather columns, you have to create two new columns!
## First: where the names of the gathered columns go (key)
## Second: where the values inside the gathered columns go (value)  
```


## Part 4 

With this new data set, calculate the average of each
`measurement` in each `year` for each different `plot_type`. Then
`spread()` them into a data set with a column for `hindfoot_length` and
`weight`. *Hint*: You only need to specify the key and value
columns for `spread()`.


```{r, answer=TRUE, purl=FALSE}
surveys_long %>%
   group_by(year, measurement, plot_type) %>%
   ## group by three variables! 
   summarize(mean_value = mean(value, na.rm=TRUE)) %>%
   ## summarize the mean of each measurement at each plot type, each year 
   spread(measurement, mean_value)
   ## spread the measurements across the columns and fill each column with the mean_value for that measurement/plot type/year
```

