---
title: "Introduction to R"
author: "Allison Theobold & The Carpentries"
output:
  pdf_document:
    fig_height: 3
    fig_width: 5
  html_document:
    fig_width: 5
  word_document:
    fig_height: 3
    fig_width: 5
---

# Introduction to Programming in `R` (30 minutes) 

The term `R` is used to refer to both the programming language as well as the 
software that interprets the scripts written using it. The learning curve may be
steeper than with other statistical software, but with `R` the results of your 
analysis or your plot does not rely on remembering what order you clicked on 
things, but instead on the written commands you generated. In `R` you will work
in scripts or with dynamic documents, with scripts within them (Rmd or Rnw files). 
Scripts may feel strange at first, but they make the steps you used in your 
analysis clear for both you and for someone who wants to give you feedback, 
further promoting the importance of reproducible science!  

# Work Flow in `R`  

Many people tend to organize their projects poorly! There are many reasons we
should *always* avoid this shamble of organization: 

* it is really hard to tell which version of your data is original and what
versions are modified  
* things get really messy because all types of files are mixed together  
* it's probably hard for you to find things and relate the correct files to 
their respective output  

# Working in `R`  

RStudio allows for you to execute commands directly from the code chunk in the 
document by using the __Ctrl + Enter__ (on Macs, __Cmd + Return__) shortcut. If
you place your cursor on the line in the code chunk that you would like to run 
and hit this shortcut, R will execute that line(s) of code for you. 
Alternatively, you can also execute code in the console (where the output of the
commands pops up). The difference between running code in the console and in the
document is that any code you execute in the console will be lost once you close
your `R` session. If you type code into the document's code chunks, it will be 
saved when you close your `R` session. Because we want to be able to go back and
re-run our code after today's workshop, it is better to type the command we want
`R` to run in the document and save it!

If R is ready to accept commands, the `R` console (in the bottom-left) will 
show a `>` prompt. When `R` receives a command (by typing, copy-pasting, or 
using the shortcut), it will execute it, and when finished will display the 
results and show the `>` symbol once again.  
If `R` is still waiting for you to provide it with additional instructions, a `+`
will appear in the console. This should tell you that you didn't finish your 
command. You could have forgotten to close your parenthesis or a quotation. If 
this happens and you are unsure of what went wrong, click inside the console and 
hit the `Esc` key. Then you can start over and figure out where you went wrong!


## Calculator  
__Practice__: Enter each of the following commands and confirm that the response
is the correct answer.  

```{r calculator, eval = FALSE}  
1 + 2 

16*9

sqrt(2)

20/5

18.5 - 7.21

3 %% 2 ## what is this doing?
```


## Creating Objects 

These operations, however, are not very interesting. To do more useful things in
`R`, we need to assign values to an object. To create an object, we tell `R` the
object's name, followed by an assignment arrow (`<-`), and finally the value of 
the object. This would look something like this:

`x <- 6`

Once we execute/run this line of code, we notice that a new object appears in 
our environment window. This window shows all of the objects that you have 
created during your `R` session. The value of `x` appears next to it, since it 
is a scalar. 

## Working with Objects 

When you assign a value to an object (like we did previously) `R` does not output
anything by default. If you enclose the code you wrote in parenthesis, then `R` 
will output the value of the object you created. 

```{r}
(x <- 6)
```

Once the object has been created, you can use it! Run the following lines of code:

```{r}
2.2 * x

4 + x 

x %% 3
```

We can also overwrite an object's value, so that it has a new value. In the code
below we give `x` a new value of 2 and use that to create a new object `y`. 

```{r}
x <- 2 

y <-  x + 6
```


**Exercise 1:** Change the value of `x` to back to 6 and see what the value of 
`y` is. Did it change from before? Is the value of `y` 8 or 12?

```{r}
# Exercise 1 code here!

```

## Working with Different Data Types 

A vector is the basic data type in `R`. A vector is a series of values, which 
can be either numbers or characters, but every entry of the vector must be the 
same data type. `R` can tell that you are building a vector when you use the 
`c()` function, which concatenates a series of entries together. 

```{r}
temps <- c(50, 55, 60, 65)
temps
```

To make a vector of characters, you are required to use quotation marks (" ") to 
indicate to `R` that the value you are using is not an object. 

```{r}
animals <- c("cat", "dog", "bird", "fish")
animals
```

Important features of a vector is the type of data they store. Run the following
lines of code and decide what type of data the vectors contain. 

```{r}
class(temps)

class(animals)
```

**Exercise 2:** Use your neighbor to create a vector that contains decimal 
valued numbers. Then check, what data type does that vector contain?

```{r}
# Exercise 2 code here!

```


Another possible data type is a logical (Boolean) value. This type of vector 
takes on values of `TRUE` and `FALSE`. But, we said that vectors could only be 
numbers or characters. If `TRUE` and `FALSE` don't have quotations around them,
then they aren't characters. So, then they must be numbers. What numbers do you 
think they are?  


```{r}

logic <- c(TRUE, FALSE, FALSE, TRUE)

class(logic)
```


**Exercise 3:** What happens when we try to mix different data types into one 
vector? Talk with your neighbor to speculate what will happen when we run each 
of the following lines of code:

```{r}
num_char <- c(1, 2, 3, "a")

num_logic <- c(1, 2, 3, FALSE)

char_logic <- c("a", "b", "c", TRUE)

guess <- c(1, 2, 3, "4")
```

In each of these vectors, the two types of data were *coerced* into a single 
data type. This happens in a hierarchy, where some data types get preference 
over others. Can we draw a diagram of the hierarchy?  


## Lists  
 
While the elements of vectors have to be of the same data type, a list is an `R`
object that allows for you to store any number of any type of object. If you 
have a vector, a matrix, a character, you can store all of them into one list 
object!

The arguments to the list function are the components of the list. Where the 
components can be characters, vectors, matrices, or other data structures. 
Here, we create a list whose components are the three vectors we've been 
working with:

```{r}
my_first_list <- list(animals, temps, logic)
my_first_list
```


__TAKE A BREAK!__

# Working with Data in `R` (30 minutes) 

## Importing Data  

* Use the __Import Dataset__ button in the __Environment__ tab  
* Choose the __From CSV__ option  
* Click on the __Browse__ button  
* Direct the computer to where you saved the BlackfootFish data file, click __open__  
* It will bring up a preview of the data  
* Click on the __Import__ button  

Notice the code that outputs in the console (the bottom left square). This is 
the code that you could have typed in the code chunk below to import the data 
yourself. Copy and paste the code that was output in the code chunk below.  

```{r data}
# copy and paste the code that was used by R to import the data 
# be careful to only copy the code that is next to the > signs! 

```

## Structure of Data  

The data we will use is organized into data tables. When you imported the 
BlackfootFish data into RStudio was saved as an object. You are able to inspect
the structure of the BlackfootFish object using functions built in to `R` 
(no packages necessary).  

Run the following code. What is output from each of the following commands?

```{r structure}
class(BlackfootFish)

dim(BlackfootFish) 
## What is the first number represent? What about the second number?

names(BlackfootFish)

str(BlackfootFish)

summary(BlackfootFish) 
## What is the data type of each variable in our dataset?
```


When we inspect dataframes, or other objects in `R`, there are some general 
functions that are useful to check the content/structure of the data. 
Here are some:

- size:  
    + `dim(data)`: rows and columns  
    + `nrow(data)`: number of rows  
    + `ncol(data)`: number of columns  

- content:  
    + head(data): first 6 rows 
    + tail(data): last 6 rows  
    + View(data): opens viewer window in separate tab  

- names:  
    + colnames(data): column names of dataframe  
    + rownames(data): row names of dataframe  

- summary of content:  
    + `str(data)`: structure of object and information about the columns  
    + `glimpse(data)`: similar information to `str`, but neater output
    + `summary(data)`: summary statistics for each column  
  

## Dataframes 

What is a dataframe? A dataframe is a type of `R` object and is the *de facto* 
structure of tabular data. You can create dataframes by hand, but most of us do 
not use `R` to input our data by hand. Instead, we import our data using `R` 
commands that read in spreadsheets (`read.csv`, `read_xls`, etc.). A dataframe 
is a set of columns, where each column is a vector. Thus, columns have the same 
data type *within* the column, but potentially different data types *across* 
columns. 

For example, the columns `trip`, `mark`, and `year` are integers (whole numbers),
`weight` and `length` are numeric (numbers with decimals), and `section` and 
`species` are characters. 


## Extracting Data  

If we were interested in accessing a specific variable in our dataset, we can 
use the `$` command. This command extracts the specified variable (on the right
of the `$` sign) from the dataset. When this is extracted, `R` views the variable
as a vector of entries, which is what the [1:18352] refers to.

```{r extract, tidy = TRUE, purl = TRUE}  

years <- BlackfootFish$year 
## extracts year from the dataset and saves it into a new variable named year

str(years) ## using the new variable (remember case matters!)

## How would you determine how long the vector is?

```

Another method for accessing data in the dataset is using matrix notation 
([row, column]). If you look to your right in the __Environment__ window, you 
notice that RStudio tells you the dimensions of the BlackfootFish data. You can 
(roughly) view the dataset as a matrix of entries, with variable names for each 
of the columns. I could instead use bracket notation to perform the same task as
above, using the following code.  

```{r bracket, purl = TRUE}  
years <- BlackfootFish[, 5] ## This takes ALL rows of data but only the fifth column
## Same as years <- BlackfootFish[1:18352, 5]

str(years)
```

## Practice:

The following is a preview of the dataframe `df`: 

```{r, echo = FALSE}
df <- data.frame(
  x = c("H", "N", "T", "W", "V"), 
  y = c("May", "Oct", "Mar", "Aug", "Feb"),
  z = c(2010, 2015, 2018, 2017, 2019)
  )

head(df)
```

**Exercise 4:** What would you input to get an output of `2015`? How did your 
neighbor do it? Was their method different?  

```{r}
# Exercise 4 code here!

```

**Exercise 5:** What would be output if you entered: `df[3, ]`? How is that 
different from entering `df[, 3]`?    

```{r}
# Exercise 5 code here!

```


**Exercise 6:** How would you pull off only columns `x` and `y`? How would you 
modify this approach to pull off columns `x` and `z`?  

```{r}
# Exercise 6 code here!

```


## Changing Data Type  

Consider the variables species and section. These variables represent a
broader class of what we call categorical variables. In `R` there are two ways 
to store this information, (1) as a series of `character` strings, or (2) as a 
`factor`. In the early days of coding in `R`, factors were more efficient than 
characters, since you only need to store the level of the factor each 
observation went with. 

While factors are still useful in today's statistical analyses and data 
visualizations, they can be tricky to deal with. When you convert a variable 
to a factor, for many operations you will get different results than for a 
character (McNamara & Horton, 2017). 

In these data section has two levels (Johnsrud and ScottyBrown) and species has
four levels (RBT, WCT, Bull, and Brown). If we want `R` to view these variables
as factors instead of characters, we need to change their data type.   


```{r factors, purl = TRUE}  

unique(BlackfootFish$species)  
## tells you the unique values of species

unique(BlackfootFish$section) 
## tells you the unique values of section  

BlackfootFish$speciesF <- as.factor(BlackfootFish$species)
## creates a new variable that is the factor version of species
BlackfootFish$sectionF <- as.factor(BlackfootFish$section)
## creates a new variable that is the factor version of section
```


There is also a function that will allow for you to specify the order of the 
levels of a factor! As we saw before, the `as.factor` function chooses the 
levels alphabetically. Suppose you would like for the species to be in the 
following order: Bull, Brown, RBT, and WCT. 

Using the `factor` function this would look like:

```{r}
BlackfootFish$speciesF <- factor(BlackfootFish$species, 
                                levels = c("Bull", "Brown", "RBT", "WCT"))
```

## Practice:  

**Exercise 7:** Year was saved as an integer data type (1989 - 2006), but it is 
a categorical variable (a factor). Write the `R` code to create a new variable 
called `yearF` that is a factor of `year` (as you did with section and species). 

```{r}
# Exercise 7 code here!

```

**Exercise 8:** Now, verify that `yearF` is viewed as a categorical variable, 
with the same levels as `year`. 
(hint: you have already used three functions that would do this for you)  

```{r}
# Exercise 8 code here!

```


The issue with factors, lies with if/when you want to change it back to a 
number. In the code below I've decided that I don't want year to be a factor 
and want to change it back to numeric. What happens when I use the 
`as.numeric()` function on the `yearF` variable?  

```{r, echo = FALSE}
BlackfootFish$yearF <- as.factor(BlackfootFish$year)
```

```{r factor_convert}

year_recover <- as.numeric(BlackfootFish$yearF)

ds <- data.frame(BlackfootFish$yearF, year_recover)
head(ds)
```

## Packages  

As we mentioned previously, `R` has many packages, which people around the world
work on to provide and maintain new software and new capabilities for `R`. 
You will slowly accumulate a number of packages that you use often for a variety
of purposes. In order to use the elements (data, functions) of the packages, you
have to first install the package (only once) and then load the package (every 
time).  

We're going to install a few packages that are often used.  

* Use the __Install__ button in the __Packages__ tab  
* Type in `devtools` and `tidyverse` into the blank line (separated by a comma)  
* Check the __Install dependencies__ box  
* Click on the __Import__ button  

There will be a large amount of output coming out of the console. This output is
`R` trying to download the package(s) you requested by contacting the mirror you
chose for it to use when downloading (I chose Northern Michigan University). 
Once the computer has downloaded the packages, it will tell you that 
"`The downloaded binary packages are in`", followed by the location of the files.  

Now that the files are downloaded, we need to load them in order to use them. 
The following code will load each package, please run it!  

```{r packages}  
library(devtools)
library(tidyverse)
```


Notice that when loading the `tidyverse` package that there is a large amount of
output. This output is telling you all of the other packages that are loaded in 
the `tidyverse` package, as well as the functions in the `tidyverse` package
that overwrite (mask) functions from base `R`.  

This is the process you go through if you ever find packages that you would 
like to use! Often packages that you install will need to be updated. To update
a package you can click on the "Tools" tab, then click on "Check for Package
Updates". This will bring up a window that will list all of the packages that 
have newer versions than what you have. Click on the packages that you wish to 
update, or click on the "Select All" button.  

## Finding Help  

One of the chief reasons for `R`'s religious following is its wonderful 
documentation. If you know a function does what you want (say find the variance),
but are not quite sure how it's spelled, what arguments it takes, or what package
it lives in, don't fret! The `?` and `help()` commands are very powerful. For 
functions, placing the `?` before the name, will tell `R` to search for that 
name in all of the functions, in all of the packages you have installed.  

* If it finds _one_ __identical match__, it will display the help file for that 
function in the Help tab in the bottom-right corner.  
* If it finds _more than one_ __identical match__, it will display the functions,
in their respective packages, that you have to choose from.  
* If it find _no_ __identical match__, it will tell you that "`No documentation 
for ____ in specified packages and libraries:,`" and suggests you use a `??` 
instead.  
    
    + A `??` in front of the function name will search __all__ of `R` for named 
    functions similar to what you typed.  
    + The output will tell you what package the function is in, as well as the 
    function's name (`package::function`).  


```{r help, eval = FALSE, purl = TRUE}    
?str
help(str)

?Levels ##incorrect function name, case sensitive
??Levels ##looks through all installed packages for a match
```

If you would like help on a particular package, say one that you just downloaded,
then you can use the same command(s) to get help.  

```{r package_help, eval = FALSE}
?tidyverse

help(tidyverse)
```

These commands will load up a help page (in RStudio) in the Help pane. Each 
help page is broken down into sections"  

* Description: An extended description of what the function does.   
* Usage: The arguments of the function and their default values.     
* Arguments: An explanation of the data each argument is expecting.  
* Details: Any important details to be aware of.    
* Value: The object the function returns.      
* See Also: Any related functions that may be useful.    
* Examples: Some examples for how to use the function.

__TAKE A BREAK!__

## Functions  

In `R` there are both functions that are built in (require no package to be 
loaded), as well as functions that are housed within specific packages. You have
already used a few built in functions to inspect the structure of the 
BlackfootFish data (`str`, `class`, `summary`). As we know, a function 
transforms an input into an output. You have to provide `R` with the inputs 
(arguments) required for the function to generate an output. The argument(s) 
inside a function happen after the `(` symbol. You know an object is a function
when it is immediately followed by a  `(` and the corresponding closing `)` 
comes after the arguments are complete. The output of a function does not have 
to be numerical and it doesn't have to be singular, it can be a set of things or
a dataset.   

Arguments describe the details of what a function is to do. Some functions take 
arguments that are specified by the user, or, if left undeclared, take on 
default values. These arguments are typically given names (as seen in the help
file), but the arguments are assumed to follow the order the function 
expects if they are not named (also stated in the help file). When naming an 
argument, the name of the argument is followed by an `=` sign and then the value
of the argument. Notice that here we are using the `=` to declare what value 
each argument is taking on, we **are not** creating a new variable with that 
value assigned to it.   

Suppose we wanted to create a vector of 10 zeros. To do this, we would use both
the `rep` function:  

```{r rep, purl = TRUE} 
# ?rep

rep(0, times = 10) ## repeating 0 three times

rep(times = 10, 0) ## switching order of arguments

rep(0, 10) ## no named arguments

rep(10, 0) ## not what we wanted!
```

Now let's look over some other functions that are often used: 


```{r functions, purl = TRUE, eval = FALSE} 

mean(BlackfootFish$weight)  
## takes a numerical input, but there are NA's in our data

mean(BlackfootFish$weight, argument here! )  
## add in the argument that removes the NA's

median(BlackfootFish$species) 
## gives an error because the input is not the correct data type  

cor(BlackfootFish$length, BlackfootFish$weight) 
## takes multiple inputs separated by a comma

## Does cor have an option to remove NA's?
```

As seen in the functions above, some functions have *optional* arguments. If 
they are not specified by the user, then they take on their default value 
(`FALSE` for `na.rm`). These options control the behavior of the functions, 
such as whether it includes/excludes NA values. 


## Cleaning Data  

In many instances, you will deal with data that are not "clean". Based on the 
output we received from the `mean()` function, we know that there are NA's in 
the BlackfootFish data, possibly across a variety of variables. Before we used 
`na.rm` as an option to remove NA's *within* a function, but the `na.omit` 
function takes a dataframe and removes any NA's from that dataset. Based on the
output below, how many rows in the BlackfootFish data have an NA present?  


```{r NA, purl = TRUE, tidy = TRUE}

dim(BlackfootFish) ## gives the dimensions of the dataset in (row, column) format  

dim(na.omit(BlackfootFish)) 
## na.omit takes dataframes, matricies, and vectors and returns the object with incomplete cases removed (NA's removed)
```


## Subsetting Data  

If we wish to remove all of the NA's from the dataset, we can simply use the 
`na.omit` command from above. We can save the new "clean" dataset under a new name
(creating a new object) or under the same name as before (replacing the old 
object with the new object).


```{r clean, tidy = TRUE}
BlackfootFish_clean <- na.omit(BlackfootFish) 
## Creates a new dataframe, where the NA's have all been removed 
```

## Data Visualization  (30 minutes)

### Scatterplots

The main purpose of the scatterplot is to show the relationship between two 
variables across several or many cases. Most often, there is a Cartesian 
coordinate system in which the x-axis represents one variable and the y-axis the
second variable.  

```{r scatter1, purl = TRUE}
#?plot()
plot(length ~ weight, 
     data = BlackfootFish_clean) 

```


```{r scatter2, purl = TRUE}
plot(length ~ weight, 
     data = BlackfootFish_clean, 
     xlab = "Weight (grams)", ## adds in axis labels 
     ylab = "Length (cm)", 
     las = 1 ## changes orientation of axis labels
     ) 
```

### Distribution

A histogram shows how many observations fall into a given range of values of a 
variable.  

```{r hist1, purl = TRUE}
hist(BlackfootFish_clean$length) 
```


```{r hist2, purl = TRUE}
hist(BlackfootFish_clean$length, 
     freq = F ## converts to a density plot (area adds to 1) 
    )
```


```{r, eval = FALSE}
## Does freq need to be named?
hist(BlackfootFish_clean$length, FALSE)
## Why is there an error about the "number of breaks"?

```


```{r hist3, purl = TRUE}
hist(BlackfootFish_clean$length, 
     freq = F, 
     xlab = "Length", ## adds x-axis lable 
     main = "Fish Lengths in Blackfoot River" ## adds title to plot
     ) 
```


```{r hist4, purl = TRUE}
hist(BlackfootFish_clean$length, 
     freq = F, 
     nclass = 50, ## changes the number of bins
     xlab = "Length", 
     main = "Fish Lengths in the Blackfoot River", 
     las = 1 ## changes orientation of axis labels
     ) 
```

### Side-by-Side Boxplots  

The familiar boxplot is effective when the objective is to compare the 
distribution of a quantitative variable across different levels of a categorical
variable. 

```{r boxplot} 
## What other options are available to add to your boxplot?
boxplot(weight ~ species, 
        data = BlackfootFish_clean)

```

### Bar Charts  

Bar charts are an effective way to compare the frequencies of levels of a 
categorical variable. 

```{r bar, purl = TRUE, tidy = TRUE, fig.height = 5}

section <- table(BlackfootFish_clean$section)
## tables the number of fish that were caught in each section

barplot(section, 
        xlab = "Section", ## adds axis labels
        ylab = "Number of Fish", 
        main = "Fish Caught by Section", ## adds title to plot
        las = 1, ## changes orientation of axis labels
        col = "blue", ## specifies the color of the bars
        ylim = c(0, 12000) ## specifies the range of y-values to plot
        )
```


## Practice 

**Exercise 9:** Using statistics or graphics, which year in our dataset had the 
most fish caught?  

```{r}
# Exercise 9 code here!

```


**Exercise 10:** Make a boxplot of the fish weights over the different years in 
the dataset.  

```{r}
# Exercise 10 code here!

```

