---
title: "Solutions to Practice Problems"
author: "Allison Theobold"
date: "1/21/2020"
output:
  pdf_document: default
  html_document: default
---

# Exercise 1 

What is the current value of `y`? 12 or 8?  

In `R`, objects are not dynamic, meaning that they do not automatically update
if you change the value of a variable they reference. Here, the value of `y` is
still 8, since it was created when `x` was 6. If you re-ran the line of code 
assigning `y` to be equal to `x + 6` after changing the value of `x` to 2, then 
`y` would be 8. 

# Exercise 2

Create a vector, named `dec`, that contains decimal valued numbers. Then check
what data type does that vector contain?


```{r}
dec <- c(2.1, 3.2, 5.5, 7.9)

class(dec)
## a numeric data type, same default storage as integer valued vectors
```


# Exercise 3

What happens when we try to mix different data types into one 
vector?


```{r}
num_char <- c(1, 2, 3, "a")

## converts everything to a character

num_logic <- c(1, 2, 3, FALSE)

## converts logical to a number (FALSE = 0, TRUE = 1)

char_logic <- c("a", "b", "c", TRUE)

## converts everything to a character

guess <- c(1, 2, 3, "4")

## converts everything to a character

```


The following is a preview of the dataframe `df`: 


```{r}
df <- data.frame(
  x = c("H", "N", "T", "W", "V"), 
  y = c("May", "Oct", "Mar", "Aug", "Feb"),
  z = c(2010, 2015, 2018, 2017, 2019)
  )

df
```


# Exercise 4

What would be output if you entered: `df[3, ]`?  

The output would be the third row of `df`, so T, Mar, 2018

# Exercise 5

What would you input to get an output of `2015`? Can you think of two ways to do
it?

You could use `df[3, 3]` to get to the element in the third row, third column. 
Or you could use `df$z[3]` to get to the third element in the vector `z`. 

# Exercise 6 

How would you pull off only columns `x` and `y`? What about pulling off only 
columns `x` and `z`?

To pull off two consecutive columns you could use: `df[1:2, ]` or 
`df[c("x", "y"), ]`. 

To pull off two non-consecutive columns you need to build a vector of columns 
you want. Like: `df[c(1, 3), ]` or `df[c("x", "z"), ]`

# Exercise 7 

How would you modify the script below, to get an output of `[1] 22 24`? 

```{r, eval = FALSE}
s <- c(22, 24, 49, 18, 1, 6)
s[1:2]
```

# Exercise 8

What would be output if you entered: `s[3, ]`?  

You would get an error that you used the "incorrect number of dimensions." 
Since `s` is a vector, it only has one dimension, so you don't need a comma!

# Exercise 9 

Year was saved as an integer data type (1989 - 2006), but we 
may want to consider it to be a categorical variable (a factor). Write the `R` 
code to create a new variable called `yearF` that is a factor of `year` 
(as you did with section and species). 

```{r}
BlackfootFish$yearF <- as.factor(BlackfootFish$year)
```


# Exercise 10 

Now, verify that `yearF` is viewed as a categorical variable, with the same 
levels as `year`. (hint: you have already used three functions that would do 
this for you)  

```{r}
levels(BlackfootFish$yearF) 
unique(BlackfootFish$year)
```

# Exercise 11 

Using statistics or graphics, which year in our dataset had the most fish 
caught?  

*Visually* 

```{r}
fish_caught <- table(BlackfootFish$year)
## tables the number of fish that were caught in each section

barplot(fish_caught, 
        xlab = "Year", ## adds axis labels to plot 
        ylab = "Number of Fish Caught", 
        main = "Fish Caught by Year", ## adds title to plot
        las = 1, ## changes orientation of axis labels
        col = "blue", ## specifies what color the bars should be
        ylim = c(0, 4000) ## specifies what range of y-values to plot
        )

```

*Numerically* 

```{r}
table(BlackfootFish$year)
```


**Exercise 12:** Make a boxplot of the fish weights over the different years in 
the dataset.  


```{r}
boxplot(weight ~ yearF, data = BlackfootFish)

## Need to use yearF, since the x-axis variable needs to be categorical!
```



