---
title: "Data Visualization with `ggplot2`"
author: "Allison Theobold"
output:
  pdf_document: default
  html_document: default
---


This reference card provides you with the recipes for exploratory graphics using **`ggplot2`**. For each recipe you will need to consider what ingredients need to be modified (added or subtracted), such as the data set (`data_set`), the variable name(s) (`variable`, `group_variable`, `x_variable`, `y_variable`), and any axis labels you wish to change.   


# Univariate Graphics  

## Bar Chart  

```{r, eval = FALSE}
ggplot(data = <data_set>, mapping = aes(x = <variable>)) + 
  geom_bar() + 
  labs(x = "Variable Name")
```

## Histogram  

```{r, eval = FALSE}
ggplot(data = <data_set>, mapping = aes(x = <variable>)) + 
  geom_histogram(binwidth = <NUMBER>, color = "COLOR") + 
  labs(x = "Variable Name (with units)")
```


## Density Plot  

```{r, eval = FALSE}
ggplot(data = <data_set>, mapping = aes(x = <variable>)) + 
  geom_density() + 
  labs(x = "Variable Name (with units)")
```

## Histogram with Density Plot  

```{r, eval = FALSE}
ggplot(data = <data_set>) + 
  geom_density(mapping = aes(x = <variable>)) +
  geom_histogram(mapping = aes(x = <variable>, y = ..density..), 
                 binwidth = <NUMBER>, color = "COLOR")
  labs(x = "Variable Name (with units)")
```

\pagebreak 

# Multivariate Graphics  

## Scatterplot  

```{r, eval = FALSE}
ggplot(data = <data_set>, mapping = aes(x = <x_variable>, y = <y_variable>)) + 
  geom_point() + 
  labs(x = "Name of x Variable (with units)", 
       y = "Name of y Variable (with units)")
```

## Scatterplot with (Linear) Smoother  

```{r, eval = FALSE}
ggplot(data = <data_set>, mapping = aes(x = <x_variable>, y = <y_variable>)) + 
  geom_point() + 
  geom_smooth(method = "lm") + 
  labs(x = "Name of x Variable (with units)", 
       y = "Name of y Variable (with units)")
```

## Boxplot  

```{r, eval = FALSE}
ggplot(data = <data_set>, mapping = aes(x = <group_variable>, y = <variable>)) + 
  geom_boxplot() + 
  coord_flip() + 
  labs(x = "Group Name", 
       y = "Variable Name (with units)")
```

## Violin Plot  

```{r, eval = FALSE}
ggplot(data = <data_set>, mapping = aes(x = <group_variable>, y = <variable>)) + 
  geom_violin() + 
  labs(x = "Group Name", 
       y = "Variable Name (with units)")
```

## Facetted Histograms  

```{r, eval = FALSE}
ggplot(data = <data_set>, mapping = aes(x = <variable>)) + 
  geom_histogram() + 
  labs(x = "Variable Name") + 
  facet_wrap(~ <group_variable>)
```

## Side-by-Side Boxplots  

```{r, eval = FALSE}
ggplot(data = <data_set>, mapping = aes(x = <group_variable>, y = <variable>)) + 
  geom_boxplot() + 
  labs(x = "Group Name", 
       y = "Variable Name (with units)")
```

