---
title: "Getting going in R and RStudio"
output:
  html_document:
    df_print: paged
  learnr::tutorial:
    progressive: yes
    allow_skip: yes
runtime: shiny_prerendered
description: |
  Learn how to set up R and RStudio on your machine. We will also demonstrate how to install R packages from CRAN, and install the tidyverse package.
---

```{r setup, include=FALSE}
library(learnr)
tutorial_options(exercise.timelimit = 60)
```

## Welcome

This is an interactive tutorial and was created by Dr. Mark Greenwood (greenwood@montana.edu) based on the [source code](https://github.com/rstudio/learnr/tree/master/inst/tutorials/ex-setup-r/ex-setup-r.Rmd) that was originally created by RStudio.

### Outline

This tutorial will help you set up your computer to use R. It is for you if you need to:

* Install R on your computer
* Install the RStudio IDE
* Install the `tidyverse` R package

You can skip this tutorial if you've already done these things in the last month. See notes below that discuss suggested updates to R and RStudio.

### Is this tutorial for you?

Do you need to work through the tutorial? Take the quiz below to find out.


```{r quiz1, echo = FALSE}
question("Check all that you have NOT done:",
  answer("installed R on my computer in the last month", message = "* Install R"),
  answer("installed the RStudio IDE", message = "* Install RStudio IDE"),
  answer("installed the tidyverse R package", message = "* Install Packages"),
  answer("None of the above. I've done them all.", correct = TRUE, message = "You can skip this tutorial!"),
  type = "multiple",
  incorrect = "This tutorial is here to help! To get set up read:"
)


```


## Install R

### How to install R

![](https://vimeo.com/203516510)

### Test your knowledge

The video was recorded with a previous version of R (3.3.2). Before moving to the next step, complete the installation on R. It is important to note that if you have previously installed R, you should complete the same steps to get an updated version of R. 


```{r quiz2, echo=FALSE}
quiz(caption = "Quiz - Install R",
  question("Is R free to download and use?",
    answer("Yes!", correct = TRUE, message = "R is free _and_ open source, which means that anyone can read, use, and copy the source code of the R language."),
    answer("No.")
  ),
  question("Where do you download R?",
    answer("At [www.rstudio.com/download](http://cloud.r-project.org)"),
    answer("At [cloud.r-project.org](http://cloud.r-project.org)", correct = TRUE, message = "You can also download R from [cran.r-project.org](http://cran.r-project.org)"),
    answer("At [www.r-project.org](www.r-project.org)", message = "Good try, but not exactly. www.r-project.org doesn't provide a download link, but it does provide a link to one of the websites above."),
    answer("At [www.r.com](http://cloud.r-project.org)"),
    allow_retry = TRUE
  ),
  question("How often should you update R?",
    answer("Everytime you use it.", message = "This will be too often unless you use R very rarely!"),
    answer("About once a year.", correct = TRUE, "A new version of R is released about once a year but sub-versions are released almost quarterly. Update your version of R if you encounter a bug that you cannot explain or it has been more than a year since you installed it."),
    answer("Never.", message = "A new version of R is released about once a year. I'll assume that you are using the newest version of R, which will be the fastest version with the fewest unexpected behaviors." )
  ),
  question("What is the current version of R?",
    answer("3.3.2", message = "It is at least version 4.2.1."),
    answer("3.6.2", message = "It is at least version 4.2.1."),
    answer("4.0.5", message = "It is at least version 4.2.1."),
    answer("4.2.1", correct = TRUE, "It was version 4.2.1 when I built this, but the current version might have a higher number (could be newer)."),
    answer("Other", message = "It was version 4.2.1 when I built this, but the current version might have a higher number (could be newer).")
  )
)

```

## Install RStudio

### How to install RStudio

RStudio is an Integrated Development Environment for R. What does that mean? Well, if you think of R as a language, which it is, you can think of RStudio as a program that helps you write and work in the language. RStudio makes programming in R much easier and I suggest that you use it!

![](https://vimeo.com/203516968)

### Test your knowledge

```{r quiz3, echo=FALSE}
quiz(caption = "Quiz - Install RStudio",
  question("What is the RStudio IDE?",
    answer("An application that makes it easier to use R.", correct = TRUE, message = "RStudio organizes your workspace and makes it easier to write, use, debug, and save R code. I highly recommend it and use it for all of my research and teaching and will support you through learning to use R."),
    answer("An application that let's you use R without writing any code.", message = "And thank goodness! Code provides a reproducible record of your work, which is essential for data science."),
    answer("A spreadsheet program like Microsoft Excel."),
    answer("Another name for R.", message = "R and RStudio are two separate things. R is a language, like English. Think of RStudio as a program that helps you use the language, kind of like how a word processing program helps you write in English."),
    allow_retry = TRUE
  ),
  question("Is the RStudio IDE free to download and use?",
    answer("Yes!", correct = TRUE, message = "Like R, RStudio is free and open-source. There do exist professional versions of RStudio, but individual users like you get access to all features you will need in the free version."),
    answer("No.", message = "Like R, RStudio is free and open-source. There do exist professional versions of RStudio, but individual users like you get access to all features you will need in the free version.")
  ),
  question("Where do you download RStudio?",
    answer("At www.rstudio.com/download", correct = TRUE, message = "For these tutorials, download and install the RStudio Desktop Open Source License."),
    answer("At [cloud.r-project.org](www.rstudio.com/download)", message = "This is where you download R, not RStudio."),
    answer("At www.r-project.org"),
    answer("At [cran.rstudio.org](http://cran.rstudio.org)"),
    allow_retry = TRUE
  ),
  question("Do you need to install R if you already have RStudio?",
    answer("Yes.", correct = TRUE),
    answer("No.", message = "R does not come with RStudio; you need to install R separately.")
  ),
  question("What is the current version of RStudio IDE?",
    answer("1.0.136", message = "It is at least version 2022.02.3."),
    answer("2022.02.3", correct = TRUE, "It was version 2022.02.3 when I built this, but the current version might have a higher number (could be newer)."),
    answer("4.2.1", message = "That was the current version of R, RStudio is at least version 2022.02.3."),
    answer("1.4", message = "It is at least version 2022.02.3."),
    answer("Other", message = "It was version 2022.02.3 when I built this, but the current version might have a higher number (could be newer).")
  )
)
```

With R and RStudio installed, you are now ready to start writing code that will allow you to load data sets and perform statistical analysis.

## Using R

### Console

It is good to start with trying out the console to do some quick code work and for things like using it as a calculator. You can see further discussion of this in [Chapter 1 of Greenwood (2020)](https://greenwood-stat.github.io/GreenwoodBookHTML/chapter1.html). 

### R Markdown

Eventually, you will want to move to using reproducible research methods via R Markdown. R Markdown provides an authoring framework for data wrangling and statistical analysis. A single R Markdown file can be used to save and execute code and 
generate high quality reports that are fully reproducible. You can go to RStudio's website for a [complete tutorial](https://rmarkdown.rstudio.com/lesson-1.html). 

That tutorial starts with this video that discusses some of what you can do with R Markdown:

![](https://vimeo.com/178485416)

We will discuss using R Markdown more in the workshop, but hopefully this gets you excited about learning to use it!


## Install Packages

### How to install R packages

![](https://vimeo.com/203516241)

### Test your knowledge

```{r names, echo = FALSE}
quiz(caption = "Quiz - Working with Packages",
  question("What command do you use to install packages?",
    answer("`library()`", message = "We will see what `library()` does later."),
    answer("`install.packages()`", correct = TRUE),
    answer("`install_packages()`", message= "Close but not quite the syntax."),
    answer("There is no command. You must visit [cran.r-project.org](http://cran.r-project.org) and download packages manually.", message = "R makes it easy to download packages. You connect to the internet and then run one of the commands above."),
    allow_retry = TRUE
  ),
  question("How often do you need to install a package on your computer?",
    answer("Every time you restart R"),
    answer("Every time you restart your computer"),
    answer("Only once. Afterwards, R can find it on your hard drive as needed.", correct = TRUE),
    answer("Never, as long as you are connected to the internet.", message = "This could be true if you are using R over a cloud service. However, if you are using R locally on your own computer, you will need to install each package that you use on your computer."),
    allow_retry = TRUE
  ),
  question("What is the tidyverse?",
    answer("A collection of packages that work well together and provide tools for common data science tasks.", correct = TRUE, message = 'You can install all of the tidyverse packages at once with `install.packages("tidyverse")`.'),
    answer("A word processing program for R"),
    answer("A code style that reads like poetry"),
    answer("An alternate universe where everyone is an R programmer."),
    allow_retry = TRUE
  )
)
```

## Wrap up (and a backup option)

Once you have completed these steps, you should be ready for the "Introduction to R" Workshop. If you had any difficulties with getting your computer set up for the workshop, please contact the workshop organizers in advance. We can either diagnose the issue via email or we can set up a time to chat and hopefully resolve the issue. 

You can also borrow a laptop from the MSU library (or maybe from a good friend) for the workshop. It needs to be a computer that is running Linux, Windows, or a somewhat recent macOS. Unfortunately, a chromebook or ipad are not sufficient for installing and running R.

One way or another, you should be set up and ready to harness the amazing capabilities of R via RStudio. Thanks for taking the time to prepare for your first steps in learning to use R. 

If you already have experience using R and RStudio and the tidyverse, the "Introduction to R" workshop may not be the best starting point for you in the R Workshops series. Feel free to email the workshop organizers if you have had some practice using RStudio in courses or other activities to discuss this. The "Introduction to R" is a very introductory workshop and more advanced workshops will follow.

## See you soon!
