---
title: "Worked Solutions to Exercises"
author: "Allison Theobold"
date: "October 14, 2019"
output: html_document
---


__Exercise 11:__ Recursive `for-loop`!  

Read in the BlackfootFish dataset and modify the code to write a `for-loop` to 
find the indices needed to sample every $7^{th}$ row from the dataset, starting 
with the $1^{st}$ row, until you've sampled 1900 rows. 

\vspace{0.5cm}

```{r e7, tidy = TRUE, eval = FALSE}
library(tidyverse)

BlackfootFish <- read_csv("data/BlackfootFish.csv")

size <- 1900

samps <- rep(NA, 1900)
## Initializing the samps vector, for storing indicies

samps[1] <- 1
## Setting first sample index to 1 (first row)

# Code snippet: samps[i] <- samps[i-1] + 7  
# The i-th sample is the i-1 sample plus 7 
# e.g. The 2nd sample is the 1st sample plus 7

for(i in 2:size){
  samps[i] <- samps[i-1] + 7
}

testing <- BlackfootFish[samps, ]
## Extracts only the rows of the dataframe in samps
training <- BlackfootFish[-samps, ]
## Excludes everything EXCEPT the rows in samps
```


__Exercise 13:__ Modify the code snippet above to instead refer to a temporary 
variable `x`.    

\vspace{0.25cm}

```{r temporary, tidy = TRUE}
# Substitute BlackfootFish$length for x everywhere! 
# By not using BlackfootFish$x, you are able to use this function with ANY dataset 

(x - min(x, na.rm = TRUE)) / 
  (max(x, na.rm = TRUE) - min(x, na.rm = TRUE))

```


__Exercise 14:__ Create another intermediate variable called `rng` that contains 
the range of the intermediate variable `x`, using the `range()` function. Make 
sure that you specify the `na.rm()` option to ignore any NAs in the input vector. 

\vspace{0.25cm}

```{r, eval = FALSE, tidy = TRUE}
# Create rng = the range of x 
rng <- range(x, na.rm = TRUE)


# Rewrite the code snippet from Excercise 13 to now 
# refer to rng instead of min(x) and max(x)
# rng[1] is the minimum 
# rng[2] is the maximum 

(x - rng[1]) / (rng[2] - rng[1])

```


__Exercise 15:__ Using the template above, write a function that rescales a 
vector to be between 0 and 1. The function should take a single argument, `x`.   

\vspace{0.25cm}

```{r function, tidy = TRUE}
# Use the function template to create a function named rescale01 that perfoms the 
# process outlined above, using the intermediate variables you previously defined

scale_norm <- function(x){
  rng <- range(x)
  (x - rng[1]) / (rng[2] - rng[1])
}

```


__Exercise 16:__ Test your function on a simple vector, with the same name as 
your function's input. What do you expect the values of the test to be after you
input it into your function? 

\vspace{0.25cm}

```{r test, tidy = TRUE}
# Test your function out on a simple vector named x
x <- c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10)

scale_norm(x)
# Looks correct! The minimum (1) becomes 0 and the maximum (10) becomes 1. 
```


__Exercise 17:__ Test your function on the `length` column of the BlackfootFish 
dataset. Inspect the rescaled values. Do they look correct?  

\vspace{0.25cm}

```{r test2, tidy = TRUE}

# Test your function out on the BlackfootFish data
scale_length <- scale_norm(BlackfootFish$length)

summary(scale_length)
# Looks right! 
```


__Exercise 18: Putting it All Together!__

```{r, eval = FALSE, tidy = TRUE}

matrix <- data.frame(l = BlackfootFish$length, w = BlackfootFish$weight)

```

\vspace{0.5cm}

Now, given the matrix of fish lengths and weights above, use the tools from the 
workshop (`for loop`, function, conditional and relational statements) to:  

* compute the condition index of each fish 
* remove the fish from the dataset whose condition index is NA or less than 2


__Condition Function__

```{r, eval = FALSE}
condition <- function(l, w){
  ## function to compute condition number of fish 
  ## data = dataframe, with columns named l & w  
  
  ## checks if weight OR length are NA before computing condition
  ifelse(is.na(w) == FALSE | is.na(l) == FALSE, 
         c <- ((w^(1/3))/l)*50, ## if both are NOT an NA, then the condition is calculated and stored in c
         c <- NA ## if either are NA, then an NA is stored in c
  )
    return(c)
    ## the value of c is returned
}
```


__For Loop Approach__ 

```{r}
## looping over the number of rows in the dataset (first entry of dim, e.g. dim[1]) 
## calculating the condition of each row 
## adding that to the condition column 

matrix$condition <- rep(NA, dim(matrix)[1])
## Creating a new column called condition

for(i in 1:dim(matrix)[1]){
  matrix$condition[i] <- condition(l = matrix$l[i], w = matrix$w[i])
}

## subseting the data to not have conditions with NA or less than 2
clean_matrix <- subset(matrix, is.na(condition) != TRUE | condition < 2)

```


__Vectorized Approach__ 

```{r}
## using the function to create a new column in the matrix dataframe
## directly using the length and weight columns in the function call = vectorized!
matrix$condition <- condition(matrix$l, matrix$w)

## subseting the data to not have conditions with NA or less than 2
clean_matrix <- subset(matrix, is.na(condition) != TRUE | condition < 2)

```


