---
title: "Challenges"
author: "Your Name Here!"
date: "10/17/2019"
output: html_document
---

> ### Challenge 1 
>
>  Using pipes, subset the `surveys` data to include animals collected before
>  1995 and retain only the columns `year`, `sex`, and `weight`.
> 
> 
```{r, answer=TRUE, eval=FALSE, purl=FALSE}
surveys %>%
     filter(year < 1995) %>%
     select(year, sex, weight)
```

> ### Challenge 2
>
>  Create a new data frame from the `surveys` data that meets the following
>  criteria: contains only the `species_id` column and a new column called
>  `hindfoot_cm` containing the `hindfoot_length` values converted to centimeters.
>  In this `hindfoot_cm` column, there are no `NA`s and all values are less
>  than 3.
>
>  **Hint**: think about how the commands should be ordered to produce this data frame!
> 
```{r, answer=TRUE, eval=FALSE, purl=FALSE}
surveys_hindfoot_cm <- surveys %>%
     filter(!is.na(hindfoot_length)) %>%
     mutate(hindfoot_cm = hindfoot_length / 10) %>%
     filter(hindfoot_cm < 3) %>%
     select(species_id, hindfoot_cm)
```



> ### Challenge 3 {.challenge}
>
> 1. How many animals were caught in each `plot_type` surveyed?
>
> ```{r, answer=TRUE, purl=FALSE}
> surveys %>%
>     count(plot_type) 
> ```
>
> 2. Use `group_by()` and `summarize()` to find the mean, min, and max hindfoot
> length for each species (using `species_id`). Also add the number of
> observations (hint: see `?n`).
>
> ```{r, answer=TRUE, purl=FALSE}
> surveys %>%
>     filter(!is.na(hindfoot_length)) %>%
>     group_by(species_id) %>%
>     summarize(
>         mean_hindfoot_length = mean(hindfoot_length),
>         min_hindfoot_length = min(hindfoot_length),
>         max_hindfoot_length = max(hindfoot_length),
>         n = n()
>     )
> ```
>
> 3. What was the heaviest animal measured in each year? Return the columns `year`,
> `genus`, `species_id`, and `weight`.
>
> ```{r, answer=TRUE, purl=FALSE}
> surveys %>%
>     filter(!is.na(weight)) %>%
>     group_by(year) %>%
>     filter(weight == max(weight)) %>%
>     select(year, genus, species, weight) %>%
>     arrange(year)
> ```


> ### Challenge 4 {.challenge}
>
> 1. Spread the `surveys` data frame with `year` as columns, `plot_id` 
>   as rows, and the
>   number of genera per plot as the values. You will need to summarize before
>   reshaping, and use the function `n_distinct()` to get the number of unique
>   genera within a particular chunk of data. It's a powerful function! See
>   `?n_distinct` for more.
> 
> ```{r, answer=TRUE, purl=FALSE}
> surveys_spread_genera <- surveys %>%
>   group_by(plot_id, year) %>%
>   summarize(n_genera = n_distinct(genus)) %>%
>   spread(year, n_genera)
> 
> head(surveys_spread_genera)
> ```
>
> 2. Now take that data frame and `gather()` it again, so each row is a unique
>    `plot_id` by `year` combination.
>
> ```{r, answer=TRUE, purl=FALSE}
> surveys_spread_genera %>%
>   gather("year", "n_genera", -plot_id)
> ```
>
> 3. The `surveys` data set has
>    two measurement columns: `hindfoot_length` and `weight`.  This makes it
>    difficult to do things like look at the relationship between mean values of
>    each measurement per year in different plot types. Let's walk through a
>    common solution for this type of problem. First, use `gather()` to create a
>     dataset where we have a key column called `measurement` and a
>    `value` column that takes on the value of either `hindfoot_length` or
>    `weight`. *Hint*: You'll need to specify which columns are being gathered.
>
> ```{r, answer=TRUE, purl=FALSE}
> surveys_long <- surveys %>%
>   gather("measurement", "value", hindfoot_length, weight)
> ```
>
> 4. With this new data set, calculate the average of each
>    `measurement` in each `year` for each different `plot_type`. Then
>    `spread()` them into a data set with a column for `hindfoot_length` and
>    `weight`. *Hint*: You only need to specify the key and value
>    columns for `spread()`.
>
> ```{r, answer=TRUE, purl=FALSE}
> surveys_long %>%
>   group_by(year, measurement, plot_type) %>%
>   summarize(mean_value = mean(value, na.rm=TRUE)) %>%
>   spread(measurement, mean_value)
> ```
